---
title: 'Final Project: CO₂ Emissions Visualisation'
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
  powerpoint_presentation: default
  word_document: default
---

# Project Overview

- **Aim:** Explore how global CO₂ emissions have evolved over time and compare per-person burdens.  
- **Dataset:** Our World In Data CO₂ (1750–present).  

- **Audience:** Educators and general public interested in climate change.

- **Tools:** tidyverse (data wrangling), ggplot2 (static plots), sf + rnaturalearth + viridis (maps).


```{r setup, message=FALSE, warning=FALSE}
# Load libraries
library(tidyverse)
library(sf)
library(rnaturalearth)
library(ggplot2)
library(viridis)
```

# Data Preparation
```{r}
# Read the OWID CO₂ dataset
co2_raw <- read_csv("owid-co2-data.csv")
```

## Keep only ISO-3 country codes
```{r}
# Filter out aggregates and non-country regions
# The OWID data uses iso_code of length 3 for countries only
co2_countries <- co2_raw %>%
  filter(str_length(iso_code) == 3)
```

```{r}
co2_countries
```

```{r}
# Top 5 countries with largest co2_per_capita
co2_countries %>%
  select(country, year, iso_code, co2_per_capita) %>%  
  group_by(iso_code) %>%                                
  slice_max(co2_per_capita, n = 1) %>%                  
  ungroup() %>%                                 
  arrange(desc(co2_per_capita))    
```

```{r}
# Top 5 countries with largest co_2 emission all time
largest_emissions <- co2_countries %>%
  select(country, year, iso_code, co2, gdp) %>%  
  group_by(iso_code) %>%                                
  slice_max(co2, n = 1) %>%                  
  ungroup() %>%                                 
  arrange(desc(co2)) %>%
  head(5)
largest_emissions
```

```{r}

```

```{r}
# CO2 consumption aggregated throughout the years
total_co2 <- co2_countries %>%
  group_by(iso_code, country) %>%          
  summarise(
    total_co2 = sum(co2, na.rm = TRUE),
    total_gdp = sum(gdp, na.rm = TRUE),
    .groups = "drop"                       
  ) %>%
  select(country, iso_code, total_co2, total_gdp) %>%
  arrange(desc(total_co2)) %>%
  head(5)

total_co2
```
```{r}
total_co2

barplot(height=total_co2$total_co2, names=total_co2$country, 
        col="#69b3a2",
        horiz=T, las=1
        )

```
## Focus countries based on total CO2 emissions
```{r}
# First, let's create a better filtered dataset name
top_co2_countries <- co2_countries %>%
  filter(iso_code %in% c("USA", "CHN", "RUS", "DEU", "GBR"))

# Create the plot with adjusted dimensions and scaling
co2_plot <- ggplot(top_co2_countries, aes(x = year)) +
  # Bar plot for GDP - adjusted scaling
  geom_bar(aes(y = gdp / max(gdp, na.rm = TRUE) * max(co2, na.rm = TRUE), 
               fill = country), 
           stat = "identity", 
           position = position_dodge(width = 0.7),  # Adjusted dodge width
           alpha = 0.6,
           width = 0.7) +  # Adjusted bar width
  
  # Line plot for CO2 emissions with points for better visibility
  geom_line(aes(y = co2, color = country), linewidth = 1) +
  geom_point(aes(y = co2, color = country), size = 2) +
  
  # Secondary axis with proper scaling
  scale_y_continuous(
    name = "CO2 Emissions (million tonnes)",
    labels = scales::comma_format(scale = 1e-6),  # Show in millions
    sec.axis = sec_axis(~ . / max(top_co2_countries$co2, na.rm = TRUE) * 
                        max(top_co2_countries$gdp, na.rm = TRUE), 
                        name = "GDP (constant $)",
                        labels = scales::dollar_format())
  ) +
  
  # Customizing colors and labels
  labs(
    title = "CO2 Emissions and GDP Over Time (Top 5 Countries)",
    x = "Year",
    color = "Country (CO2)",
    fill = "Country (GDP)"
  ) +
  
  # Theme improvements
  theme_minimal(base_size = 20) +  # Increased base font size
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.title = element_text(size = 20),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 24),
    axis.title.y = element_text(margin = margin(r = 15), size = 24),
    axis.title.y.right = element_text(margin = margin(l = 15)),
    plot.margin = margin(1, 1, 1, 1, "cm"),  # Added margins
    plot.title = element_text(hjust = 0.5, face = "bold", size = 24)
  ) +
  
  # Color scales
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1")

# Display the plot with appropriate dimensions
options(repr.plot.width = 45, repr.plot.height = 10)  # For R Notebook
co2_plot

# If saving to a file:
ggsave("co2_gdp_plot.png", co2_plot, width = 45, height = 20, dpi = 300)

```


```{r}
largest_co25 <- co2_countries %>%
  filter(iso_code %in% c("JPN", "DEU"))

  #filter(iso_code %in% c("USA", "CHN", "RUS", "JPN", "DEU"))


ggplot(largest_co25, aes(x = year)) +
  # Bar plot for GDP (with transparency using alpha)
  geom_bar(aes(y = gdp / max(gdp) * max(co2), fill = country), 
           stat = "identity", 
           position = "dodge", 
           alpha = 0.6) +  # alpha controls transparency
  
  # Line plot for CO2 emissions
  geom_line(aes(y = co2, color = country), linewidth = 1) +
  
  # Secondary axis for GDP (scaled to match CO2 range)
  scale_y_continuous(
    name = "CO2 Emissions (tonnes)",
    sec.axis = sec_axis(~ . / max(largest_co25$co2) * max(largest_co25$gdp), 
                        name = "GDP (constant $)")
  ) +
  
  # Customizing colors and labels
  labs(
    title = "CO2 Emissions and GDP Over Time Japan and Germany",
    x = "Year",
    color = "Country (CO₂)",
    fill = "Country (GDP)"
  ) +
  
  # Theme improvements
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
## Focus on five countries
```{r}
# Pick a representative set of 5 countries:
# China, United States, India, Germany, Nigeria
focus_countries <- c("CHN", "USA", "IND", "DEU", "NGA")

co25 <- co2_countries %>%
  filter(iso_code %in% focus_countries)
```

## Restrict to years 1960–2021
```{r}
co2_trim <- co25 %>%
  filter(year >= 1960, year <= 2021)
```

## Drop missing
```{r}
# 5. Drop rows with missing total or per-capita emissions
co2_clean <- co2_trim %>%
  filter(!is.na(co2), !is.na(co2_per_capita))
```

## Summarise latest year
```{r}
latest <- max(co2_clean$year)
summary_latest <- co2_clean %>%
  filter(year == latest) %>%
  select(country, co2, co2_per_capita) %>%
  arrange(desc(co2))

# Show the result
summary_latest
```

```{r}
largest_co25
```

# Results
## Figure 1. Total Annual CO₂ Emissions, 1960–2021
```{r viz-line, fig.cap="Total CO₂ emissions (1960–2021) for five countries", warning=FALSE}
ggplot(co2_clean, aes(x = year, y = co2, color = country)) +
  geom_line(size = 1) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title    = expression("Total Annual CO"[2]~" Emissions, 1960–2021"),
    subtitle = "Five illustrative countries",
    x        = "Year",
    y        = expression("CO"[2]~" Emissions (million tonnes)"),
    color    = "Country"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```



## Figure 2. Per-Capita CO₂ Emissions in 2021

```{r viz-bar, fig.cap="Per-capita CO₂ Emissions in 2021"}
percap2021 <- co2_clean %>% filter(year == 2021)

ggplot(percap2021, aes(
    x = reorder(country, co2_per_capita),
    y = co2_per_capita,
    fill = country
  )) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = bquote("Per-capita CO"[2]~" Emissions in"~.(max(percap2021$year))),
    x     = NULL,
    y     = expression("CO"[2]~" per person (tonnes)")
  ) +
  theme_minimal(base_size = 14)
```



## Figure 3. Global Per-Capita CO₂ Emissions, 2021 (Top 5 Labeled)
```{r choropleth-full, message=FALSE, warning=FALSE, fig.cap="Global Per-Capita CO₂ Emissions, 2021"}
# World map
world <- ne_countries(scale = "medium", returnclass = "sf") %>%
  select(iso_a3, geometry)

# 2021 per-capita
percap_all2021 <- co2_countries %>% filter(year == 2021) %>%
  select(iso_code, co2_per_capita)

# Join
world_percap <- left_join(world, percap_all2021,
                          by = c("iso_a3" = "iso_code"))

# Plot
ggplot(world_percap) +
  geom_sf(aes(fill = co2_per_capita), color = NA) +
  scale_fill_viridis_c(
    option   = "plasma",
    na.value = "grey90",
    name     = expression("t CO"[2]~"/ person")
  ) +
  labs(
    title    = expression("Global Per-Capita CO"[2]~" Emissions, 2021"),
    subtitle = "Data source: Our World in Data",
    caption  = "Grey = missing data"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = c(0.15, 0.25),
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(size = 12),
    plot.caption    = element_text(size = 9)
  )
```




